Social Attention for Autonomous Decision-Making in Dense Traffic

### 2019-09-11 main.tex
Update manuscript: agents

This observation has led the community to turn to learning-based methods, which bear the promise of leveraging data to automatically learn a complex driving policy. In the imitation learning approach, a policy can be trained in a supervised manner to imitate human driving decisions \citep[e.g.][]{Pomerleau1989, Ross2011, Bojarski2016, Xu2016, Eraqi2017, Codevilla2017, Rehder2017c, Rezagholiradeh2018, Rhinehart2018, Bansal2018, Rhinehart2019}. Because the cost of human driving data collection at large scale can be prohibitive, another promising approach is train a policy in simulation using reinforcement learning \citep[e.g.][]{Cardamone2009, Ross2011, Mukadam2017, Chen2017, Isele_2018, Ha2018, Kendall2019}.

@@

@@

As $Q^*$ is a fixed-point of the Bellman Operator $\cT$ -- which is a contraction --, it can be computed by applying $\cT$ iteratively in a fixed-point iteration fashion. The \emph{Q-learning} algorithm \citep{Watkins} follows this procedure by applying a sampling version $\cT$ to a batch of collected experience. When dealing with a continuous state space $S$, we need to employ function approximation in order to generalise to nearby states. The \emph{Deep Q-Network} (DQN) algorithm \citep{Mnih} implements this idea by using a neural network model to represent the action-value function $Q$.

@@

\begin{equation*}
s = \left\lbrace s_i \right\rbrace_{i \in [0, N]}\qquad
\text{where}\qquad
s_i = \begin{bmatrix}
x_i & y_i & v^x_i & v^y_i & \cos\psi_i & \sin \psi_i
\end{bmatrix}^T
\end{equation*}

@@

This encoding is efficient in the sense that it uses the smallest quantity of information necessary to represent the scene. However, it lacks two important properties. First, its size varies with the number of vehicles which can be problematic for the sake of function approximation which often expects constant-sized inputs, but also for concerns of varying computational complexity. Second, we expect a driving policy $\pi$ to be \emph{permutation invariant}, i.e. not to be dependent on the order in which other traffic participants are listed. Ideally, this property should be enforced and not approximated by relying on the coverage of the $N!$ possible permutations $\tau$ of any given traffic state. Formally, we require that:
\begin{equation*}
\pi(\cdot|(s_0, s_1,\dotsc,s_N)) = \pi(\cdot|(s_0, s_{\tau(1)},\dotsc,s_{\tau(N)})) \quad\quad\quad \forall\tau \in \mathfrak{S}_N
\end{equation*}

@@

\section{Model Architecture}

@@

The proposed architecture is presented in \autoref{fig:architecture}. It is used to represent the $Q$-function that will be optimized by the DQN algorithm. It is composed of a first linear encoding layer whose weights are shared between all vehicles. At that point, the embeddings only contain individual features of size $d_x$. They are then fed to an ego-attention layer, composed of several heads stacked together. Such an attention head is illustrated in \autoref{fig:ego-attention}, and works in the following way: in order to select a subset of vehicles depending on the context, the ego-vehicle  first emits a query $Q = [q_0]\in\Real^{1 \times d_k}$, computed with a linear projection $L_q\in\Real^{d_x \times d_k}$ of its embedding. This query is then compared to a set of keys $K = [k_0, \dots, k_N]\in\Real^{N \times d_k}$ containing descriptive features $k_i$ for each vehicle, again computed with a shared linear projection $L_k\in\Real^{d_x \times d_k}$. The similarity between the query $q_0$ and any key $k_i$ is assessed by their dot product $q_0 k_i^T$. These similarities are then scaled by the inverse-square-root-dimension $1/\sqrt{d_k}$\footnote{This scaling is due to the fact that the dot-product of two independent random vectors with mean 0,  variance 1, and dimension $d_k$, is a random variable with mean 0 and variance $d_k$} and normalized with a softmax function $\sigma$ across vehicles. We obtain a stochastic matrix called the \emph{attention matrix}, which is finally used to gather a set of output value $V = [v_0, \dots, v_N]$, where each value $v_i$ is a feature computed with a shared linear projection $L_v\in\Real^{d_x \times d_v}$. Overall, the ego-attention computation for each head can be written as:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq:selfattention}
\end{equation}
The outputs from all heads are finally combined with a linear layer, and the resulting tensor is then added to the ego encoding as in residual networks. We can easily see that the result is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows in keys $K$ and values $V$ in \eqref{eq:selfattention} but will keep their correspondence. The final result is a dot product of values and key-similarities, which is independent of the ordering.

@@

	\draw (DEC1.east) -- (Y1.west);
	\end{tikzpicture}
	\caption{Block diagram of our model architecture.}
	\label{fig:architecture}

@@

	\end{tikzpicture}
	\caption{Architecture of an ego-attention head.
		The blocks $L_{q}$, $L_{k}$, $L_{v}$ are linear layers. The keys $K$ and values $V$ are concatenated from all vehicles, while the query $Q$ is only produced by the ego-vehicle.}
	\label{fig:ego-attention}
\end{figure}

@@

TODO, see \autoref{tab:agents}.

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPC & \MLPG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [?, ?, 7] & [1-15, 7] \\
			Layers sizes & [128, 128] &  [64, 64] & \makecell[l]{Encoder: [64, 64] \\Attention: 2 heads, $d=32$ \\ Decoder: [64, 64]} \\
			Number of parameters & 3.0e4 & ? & 3.4e4 \\
			Variable input size & No & No & Yes  \\
			Permutation invariant & No & Yes & Yes \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}

@@

We show in \autoref{fig:results} the evolution of the total reward, episode length and average velocity.
The \MLPC agent learns to accelerate to earn short-term rewards, as shown by its high average velocity , but fails to exploit the information of other vehicles and crashes often, leading to shorter episodes = risky policy, worst performing.
Conversely, the \MLPG benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions (increased duration). However, it only proceeds when the intersection has been cleared (low mean velocity), resulting in an overly cautious policy, a common trait known as the freezing robot problem. Slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection but also when it can exploit the gaps in the traffic and take way to vehicles that are far/slow enough. The overall resulting behaviour is qualitatively more nuanced and human-like.

>>>>

This observation has led the community to turn to learning-based methods, which bear the promise of leveraging data to automatically learn a complex driving policy. In the imitation learning approach, a policy can be trained in a supervised manner to imitate human driving decisions \citep[e.g.][]{Pomerleau1989, Ross2011, Bojarski2016, Xu2016, Eraqi2017, Codevilla2017, Rehder2017c, Rezagholiradeh2018, Rhinehart2018, Bansal2018, Rhinehart2019}. Because the cost of human driving data collection at large scale can be prohibitive, another promising approach is train a policy in simulation using reinforcement learning \citep[e.g.][]{Cardamone2009, Ross2011, Mukadam2017, Chen2017, Isele2017, Isele2018, Ha2018, Kendall2019}.

@@

\label{sec:background}

@@

As $Q^*$ is a fixed-point of the Bellman Operator $\cT$ -- which is a contraction --, it can be computed by applying $\cT$ iteratively in a fixed-point iteration fashion. The \emph{Q-learning} algorithm \citep{Watkins} follows this procedure by applying a sampling version $\cT$ to a batch of collected experience. When dealing with a continuous state space $S$, we need to employ function approximation in order to generalise to nearby states. The \emph{Deep Q-Network} (DQN) algorithm \citep{Mnih2015} implements this idea by using a neural network model to represent the action-value function $Q$.

@@

\begin{equation}
s = \left\lbrace s_i \right\rbrace_{i \in [0, N]}\qquad
\text{where}\qquad
s_i = \begin{bmatrix}
x_i & y_i & v^x_i & v^y_i & \cos\psi_i & \sin \psi_i
\end{bmatrix}^T
\label{eq:coordinates}
\end{equation}

@@

This encoding is efficient in the sense that it uses the smallest quantity of information necessary to represent the scene. However, it lacks two important properties. First, its size varies with the number of vehicles which can be problematic for the sake of function approximation which often expects constant-sized inputs. Second, we expect a driving policy $\pi$ to be \emph{permutation invariant}, i.e. not to be dependent on the order in which other traffic participants are listed. Ideally, this property should be enforced and not approximated by relying on the coverage of the $N!$ possible permutations $\tau$ of any given traffic state. Formally, we require that:
\begin{equation*}
\pi(\cdot|(s_0, s_1,\dotsc,s_N)) = \pi(\cdot|(s_0, s_{\tau(1)},\dotsc,s_{\tau(N)})) \quad\quad\quad \forall\tau \in \mathfrak{S}_N
\end{equation*}

@@

\section{Model Architecture}
\label{sec:architecture}

@@

The proposed architecture is presented in \autoref{fig:architecture}. It is used to represent the $Q$-function that will be optimized by the DQN algorithm. It is composed of a first linear encoding layer whose weights are shared between all vehicles. At that point, the embeddings only contain individual features of size $d_x$. They are then fed to an ego-attention layer, composed of several heads stacked together. Such an attention head is illustrated in \autoref{fig:ego-attention}, and works in the following way: in order to select a subset of vehicles depending on the context, the ego-vehicle  first emits a query $Q = [q_0]\in\Real^{1 \times d_k}$, computed with a linear projection $L_q\in\Real^{d_x \times d_k}$ of its embedding. This query is then compared to a set of keys $K = [k_0, \dots, k_N]\in\Real^{N \times d_k}$ containing descriptive features $k_i$ for each vehicle, again computed with a shared linear projection $L_k\in\Real^{d_x \times d_k}$. The similarity between the query $q_0$ and any key $k_i$ is assessed by their dot product $q_0 k_i^T$. These similarities are then scaled by the inverse-square-root-dimension $1/\sqrt{d_k}$\footnote{This scaling is due to the fact that the dot-product of two independent random vectors with mean 0,  variance 1, and dimension $d_k$, is a random variable with mean 0 and variance $d_k$} and normalized with a softmax function $\sigma$ across vehicles. We obtain a stochastic matrix called the \emph{attention matrix}, which is finally used to gather a set of output value $V = [v_0, \dots, v_N]$, where each value $v_i$ is a feature computed with a shared linear projection $L_v\in\Real^{d_x \times d_v}$. Overall, the attention computation for each head can be written as:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq:selfattention}
\end{equation}
The outputs from all heads are finally combined with a linear layer, and the resulting tensor is then added to the ego encoding as in residual networks. We can easily see that this process is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows in keys $K$ and values $V$ in \eqref{eq:selfattention} but will keep their correspondence. The final result is a dot product of values and key-similarities, which is independent of the ordering.

@@

	\draw (DEC1.east) -- (Y1.west);
	\end{tikzpicture}
	\caption{Block diagram of our model architecture. It is composed of several linear identical encoders, a stack of ego-attention heads, and a linear decoder.}
	\label{fig:architecture}

@@

	\draw[draw=black, dashed] (1.75cm, -5.5cm) rectangle (9.5cm,1.5cm);
	\end{tikzpicture}
	\caption{Architecture of an ego-attention head.
		The blocks $L_{q}$, $L_{k}$, $L_{v}$ are linear layers. The keys $K$ and values $V$ are concatenated from all vehicles, while the query $Q$ is only produced by the ego-vehicle.}
	\label{fig:ego-attention}
\end{figure}

@@

We evaluate three different agents, whose characteristics are summarized in \autoref{tab:agents}.
\begin{itemize}
	\item \MLPC: a \emph{list of features} state representation is used, as described in Section \ref{sec:background}. The model is a simple multi-layer perceptron (MLP). Because this architecture requires a fixed-size input, we use zero-padding to fill the input tensor up to a maximum number $N=14$ of observed vehicles, and add an additional \emph{presence} feature to the coordinates described in \eqref{eq:coordinates} so as to identify active rows.
	\item \MLPG: a \emph{spatial grid} representation is used, as described in Section \ref{sec:background}, with a $11 \times 11$ grid where each cell represents a $5\text{m}\times 5$m square. The model is a multi-layer perceptron.
	\item \EgoAtt: a \emph{list of features} state representation is used along with the Ego-Attention architecture described in Section \ref{sec:architecture}. As this model supports varying-size inputs, zero-padding is not required.
\end{itemize}

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPC & \MLPG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [11, 11, 5] & [1-15, 7] \\
			Layers sizes & [128, 128] &  [48, 48] & \makecell[tl]{Encoder: [64, 64] \\Attention: 2 heads, $d_k=32$ \\ Decoder: [64, 64]} \\
			Number of parameters & 3.0e4 & 3.2e4 & 3.4e4 \\
			Variable input size & No & No &  {Yes}  \\
			Permutation invariant & No & {Yes} &  {Yes} \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}

These agents are all trained with the DQN algorithm using the same hyperparameters, and their architectures are scaled to admit about the same number of trainable parameters for fair comparison.

@@

We plot in \autoref{fig:results} the evolution of the total reward, episode length and average velocity during training, over 4000 episodes and repeated across 120 random seeds.
The \MLPC agent learns to accelerate to earn short-term rewards, as shown by its high average velocity, but fails to exploit the information of other vehicles and crashes often, leading to shorter episodes. In short, we obtain a risky and blind policy that is the worst performing.
Conversely, the \MLPG architecture benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions, as we can see from its higher episode length. However, it only proceeds when the intersection has been fully cleared, as reflected by its low average velocity. This results in an overly cautious policy -- a common trait colloquially known as the \emph{freezing robot problem} \citep{Trautman2010} -- but a slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection but also when it can exploit the gaps in the traffic and take way to vehicles that are far/slow enough. The overall resulting behaviour is qualitatively more nuanced and human-like.
